---
title: "Borrowing and Yields in the European Union"
subtitle: "An Automatically Updated Overview of the EU's Finances"
author: Dr Michal Ovádek
output: 
  html_document:
    df_print: paged
    number_sections: true
    toc: true
    toc_float: true
    theme: united
    highlight: pygments
---

<style>
body {
text-align: justify}
</style>

<style type="text/css">
  body{
  font-size: 13pt;
}
</style>

<style>
.tocify-extend-page {
  height: 30px !important;
}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      error = FALSE,
                      fig.width = 9,
                      fig.height = 6)
```


```{r initialdata, include=FALSE}

# packages
library(ggplot2)
library(ggforce)
library(ggiraph)
library(dplyr)
library(purrr)
library(tidyr)
library(stringr)
library(forcats)
library(rmarkdown)
library(modelsummary)
library(DT)
library(lubridate)
library(gdtools)
library(gfonts)

# seed
set.seed(35239)

# fonts
register_gfont("Open Sans")
addGFontHtmlDependency(family = "Open Sans")

# current date
date_now <- Sys.Date()
date_now_f <- str_remove(format(date_now, "%d %B %Y"), "^0")
day_now <- as.integer(str_sub(date_now, 9, 10))
year_now <- as.integer(str_sub(date_now, 1, 4))

# citation
cit_page <- paste("Michal Ovádek, '",rmarkdown::metadata$title,"', available at https://michalovadek.github.io/eufinance/, accessed on ", date_now_f,
                  sep = "")

#### download data ####

# ecb api url
ecb_api <- "https://data-api.ecb.europa.eu/service/data/"

# time period argument
"startPeriod=2004-09-06&endPeriod=2023-12-31"

# Euro area bond yields, spot rates
yields_l <- list()

# 1 year, 5 year, 10 year maturities
for (Y in c(1, 5, 10)){
  
  # construct URL
  url = str_c(
  ecb_api,
  "YC/B.U2.EUR.4F.G_N_A.SV_C_YM.SR_",
  Y,
  "Y?startPeriod=2004-09-06&",
  "endPeriod=",
  as.character(date_now),
  "&format=csvdata")
  
  yields_l[[as.character(Y)]] <- read.csv(url)
  
}

# consolidate euro area yields
yields_euroarea <- bind_rows(yields_l) |> 
  select(DATA_TYPE_FM, TIME_PERIOD, OBS_VALUE, TITLE) |> 
  as_tibble()

```

<br>

The European Union (EU) is not only a single market but also an economic and monetary union. Most EU members have a common currency -- the euro -- which entails a supranational monetary policy. The European Central Bank (ECB) manages the euro together with Euro Area central banks.

This automatically updated document provides a real-time overview of several indicators that are central to economic and monetary policy in the EU. All data comes from the [ECB API](https://data.ecb.europa.eu/help/api/) and Eurostat. The document is current as of ```r date_now_f```.

# Euro Area Government Bond Yields

Government borrowing is normally carried out through bond sales. Bonds are financial instruments that entitle the buyer (investor) to an interest rate (coupon rate) and the return of the principal on expiry (maturity date) in exchange for lending the government money at the outset. Like other securities, bonds are traded on secondary markets and their value and yields are therefore subject to fluctuation.

```{r euroareabondyields}
# iplot
iplot_yields_euro <- yields_euroarea |> 
  mutate(tooltip = case_when(
     DATA_TYPE_FM == "SR_1Y" ~ "1-year maturity",
     DATA_TYPE_FM == "SR_5Y" ~ "5-year maturity",
     DATA_TYPE_FM == "SR_10Y" ~ "10-year maturity",
     T ~ NA_character_
  )) |> 
  ggplot(aes(x = as.Date(TIME_PERIOD), y = OBS_VALUE)) +
  geom_hline(yintercept = 0, lty = 1, color = "grey83") +
  geom_line_interactive(aes(color = DATA_TYPE_FM,
                            tooltip = tooltip,
                            data_id = DATA_TYPE_FM), 
                        linewidth = 0.23,
                        show.legend = FALSE) + 
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_color_manual_interactive(values = c("#98EDDF","#2D5C54","#5AB8A8")) +
  theme_minimal(base_family = "Open Sans") +
  theme(panel.grid = element_line(color = "grey96", linewidth = 0.2),
        strip.background = element_rect(fill = "grey95", color = "grey92"),
        plot.title = element_text(face = "bold")) +
  facet_zoom(xy = between(as.Date(TIME_PERIOD), date_now-365, date_now),
             horizontal = FALSE,
             zoom.size = 0.5) +
  labs(x = NULL, y = "Yield",
       title = "Euro Area Government Bond Yields",
       subtitle = "Daily spot rates, triple A issuers only")

# interactive
girafe(ggobj = iplot_yields_euro,
       fonts = list(sans = "Open Sans"),
       #width_svg = 12,
       #height_svg = 8,
       options = list(opts_sizing(rescale = TRUE),
                      opts_toolbar(saveaspng = FALSE),
                      opts_tooltip(css = "background-color:gray;color:white;font-style:italic;padding:9px;border-radius:5px;font-size:15px;",
                                   use_fill = TRUE),
                      opts_hover_inv(css = "opacity:0.2;"),
                      opts_hover(css = ""))
)
```

Typically, lending money for longer periods of time is associated with higher interest rates. When shorter-term yields overtake longer-term yields, we speak of an inverted yield curve. Yield curve inversion is historically predictive of recessions.

```{r eurostatbonds}

# eurostat
library(eurostat)

# get long term bond yields
gov_b10yields <- get_eurostat("teimf050")

# prep data
pdata_govb10y <- gov_b10yields |> 
  filter(!geo %in% c("EA","EU27_2020")) |> 
  left_join(bind_rows(eurostat::eu_countries, tibble(code = "UK", name = "UK")), 
            by = c("geo"="code")) |> 
  mutate(month = str_sub(as.character(TIME_PERIOD), 1, 7),
         tooltip = str_c("Yield: ", values, " (", month, ")"),
         id = str_c(geo, month))

# current max and min
cur_max_yld <- pdata_govb10y |> 
  arrange(desc(TIME_PERIOD)) |> 
  group_by(name) |> 
  slice(1) |> 
  ungroup() |> 
  slice_max(values)

cur_min_yld <- pdata_govb10y |> 
  arrange(desc(TIME_PERIOD)) |> 
  group_by(name) |> 
  slice(1) |> 
  ungroup() |> 
  slice_min(values)

# EA and non-EA average
non_euro_ylds_avg <- pdata_govb10y |> 
  filter(geo %in% c("CZ","PL","HU","RO","BG","SE","DK")) |> 
  summarise(mean = round(mean(values), 2))

euro_ylds_avg <- filter(gov_b10yields, geo == "EA") |> 
  summarise(mean = round(mean(values), 2))

# iplot
iplot_govylds <- pdata_govb10y |> 
  ggplot() +
  geom_line(data = filter(gov_b10yields, geo == "EA") |> select(-geo),
            aes(x = as.Date(TIME_PERIOD), y = values),
            lty = 2,
            color = "#284ec9") +
  geom_line_interactive(aes(x = as.Date(TIME_PERIOD), y = values),
                        color = "grey60",
                        linewidth = 0.4) +
  geom_point_interactive(aes(x = as.Date(TIME_PERIOD), y = values, 
                             tooltip = tooltip, data_id = id), 
                         size = 0.4,
                         color = "grey40") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "month", labels = function(z) gsub("^0", "", strftime(z, "%m"))) +
  facet_wrap(~name) +
  theme_minimal(base_family = "sans") +
  theme(panel.grid = element_line(color = "grey96", linewidth = 0.2),
        strip.background = element_blank(),
        strip.text = element_text(hjust = 1, face = "bold"),
        axis.text.x = element_text(size = 6),
        plot.title = element_text(face = "bold")) +
  labs(x = NULL, y = "Yield",
       title = "Government Bond Yields in the EU and UK",
       subtitle = "10-year maturity, monthly data",
       caption = "Dashed line represents the Euro Area average")

# interactive
girafe(ggobj = iplot_govylds,
       fonts = list(sans = "Open Sans"),
       #width_svg = 12,
       #height_svg = 8,
       options = list(opts_sizing(rescale = TRUE),
                      opts_toolbar(saveaspng = FALSE),
                      opts_tooltip(css = "background-color:gray;color:white;font-style:italic;padding:9px;border-radius:5px;font-size:15px;",
                                   use_fill = TRUE),
                      opts_hover_inv(css = "opacity:0.2;"),
                      opts_hover(css = ""))
)

```

As most government borrowing in the EU takes place at the national level, bond yields vary not only over time but also systematically by country. At the moment, yields are highest on bonds issued by ```r cur_max_yld$name```. Last month they traded at a yield of ```r cur_max_yld$values```%. In contrast, ```r cur_min_yld$name```'s bonds had the lowest yield at ```r cur_min_yld$values```%.

Bond yields are indicative of the cost of government borrowing and as such play a major role in debates about the benefits of adopting the euro as a currency. Over the last year, 10-year bonds in the Euro Area had an average yield of ```r euro_ylds_avg$mean```%, while EU members that do not use the euro averaged a yield of ```r non_euro_ylds_avg$mean```%. This simplistic comparison is confounded by the uneven composition of the two groups in terms of macroeconomic stability, but it can be instructive to compare relatively similar countries in different regions (e.g. Czechia and Slovakia or Finland and Sweden).

# Euro Area Key Interest Rates

The ECB's primary mandate is price stability. Its long-term inflation target is 2% (year-on-year).

```{r eurokeyinterest}

## key ECB interest rate
# construct URL
url = str_c(
  ecb_api,
  "FM/B.U2.EUR.4F.KR.MRR_FR.LEV",
  "?startPeriod=1998-01-01&",
  "endPeriod=",
  as.character(date_now),
  "&format=csvdata"
)

# main refinancing operations
key_interest_rate <- read.csv(url) |> 
  select(TIME_PERIOD, OBS_VALUE, TITLE)

## inflation
# construct URL
url = str_c(
  ecb_api,
  "ICP/M.U2.N.000000.4.ANR",
  "?startPeriod=1998-01-01&",
  "endPeriod=",
  as.character(date_now),
  "&format=csvdata"
)

# year-on-year inflation 
inflation_rate <- read.csv(url) |> 
  select(TIME_PERIOD, OBS_VALUE, TITLE) |> 
  mutate(date = as.character(lubridate::ym(TIME_PERIOD)))

# days
days <- seq.Date(from = as.Date("1998-01-01"), to = date_now, by = "day")

# interest rate and inflation
rates <- tibble(date = as.character(days)) |> 
  left_join(inflation_rate |> select(-TITLE, -TIME_PERIOD)) |> 
  fill(OBS_VALUE) |> 
  rename(inflation = OBS_VALUE) |> 
  left_join(key_interest_rate |> select(-TITLE), by = c("date" = "TIME_PERIOD")) |> 
  fill(OBS_VALUE) |> 
  rename(ecb_rate = OBS_VALUE)

# iplot
iplot_rates <- rates |> 
  pivot_longer(cols = -date) |> 
  drop_na() |> 
  mutate(tooltip = ifelse(name == "inflation", "Harmonised index of consumer prices, year-on-year change", "Interest rate on main refinancing operations")) |> 
  ggplot(aes(x = as.Date(date), y = value)) +
  geom_hline(yintercept = 0, lty = 1, color = "grey83") +
  geom_hline(yintercept = 2, lty = 2, color = "grey40") +
  geom_step_interactive(aes(color = name, tooltip = tooltip, data_id = name),
                        linewidth = 0.29,
                        show.legend = FALSE) +
  annotate(geom = "text", label = "2% inflation target", x = as.Date("1997-11-01"), y = 2.4, size = 2.1) +
  #scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  scale_y_continuous(labels = function(x) paste0(x, "%")) +
  scale_color_manual_interactive(values = c("#CDC7FD", "#758055")) +
  theme_minimal(base_family = "sans") +
  theme(panel.grid = element_line(color = "grey96", linewidth = 0.2),
        strip.background = element_rect(fill = "grey95", color = "grey92"),
        plot.title = element_text(face = "bold")) +
  facet_zoom(x = between(as.Date(date), date_now-365, date_now),
             horizontal = FALSE,
             zoom.size = 0.5) +
  labs(x = NULL, y = "Rate",
       title = "Key Interest Rate and Inflation",
       subtitle = "Main refinancing operations and index of consumer prices")

girafe(ggobj = iplot_rates,
       fonts = list(sans = "Open Sans"),
       #width_svg = 12,
       #height_svg = 8,
       options = list(opts_sizing(rescale = TRUE),
                      opts_toolbar(saveaspng = FALSE),
                      opts_tooltip(css = "background-color:gray;color:white;font-style:italic;padding:9px;border-radius:5px;font-size:15px;",
                                   use_fill = TRUE),
                      opts_hover_inv(css = "opacity:0.1;"),
                      opts_hover(css = ""))
)

```

In times of rising inflation, the ECB's main countervailing instrument is increasing key interest rates.

# Cite

Cite this document as `r cit_page`.
